Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-01-08 analysis/paper.qmd
Add crop progenitors discussion

* [x] Stop modelling
- [ ] First draft:
  - [x] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [ ] Discussion
    - [x] General trends
    - [ ] Case studies
    - [x] Comparison with archaeobot
  - [ ] Conclusion
- [ ] Figures
  - [x] fig-region
  - [x] fig-occ
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
  - [ ] tbl-model-metrics
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
- [x] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Final proofread

@@

<!-- TODO: more citations? -->
Ecological niche modelling (ENM) or species distribution modelling (SDM) is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors.
Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
This model can then be used to predict the range of the organism in question either in the same or a different environment.
@CITE suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.

@@

The major practical limitation of the hindcasting approach is that it relies on spatially explicit, high resolution palaeoenvironmental surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most applications, which is perhaps why only a minority of studies use it [cf. @YaworskyEtAl2023].
In this study, we are able to take advantage of the increasing availability of high resolution, global palaeoclimate data derived from simulation experiments with general circulation models of climate [@BrownEtAl2018; @BrownEtAl2020; @KargerEtAl2023].
<!-- TODO: mention that GBIF, CHELSA, SoilGrids, WorldClim, DEMs, etc. are also great? -->

@@

@@

```{r niche-predict}
# NOTE: high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  flora <- flora |>
    mutate(
      pred = map(niche, collect_predictions),
      palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = environment, .progress = TRUE),
      # Binarize
      palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
      palaeodist = map(palaeodist, mutate, absent = prediction.FALSE. > 0.5)
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))
```

@@

```{r fig-palaeodist-sum}
#| fig-cap: Predicted species richness (sum of predicted ranges) by period
# TODO: expand period facet labels
ggplot() +
  facet_wrap(vars(factor(period, levels = climate_periods$code, labels = climate_periods$label))) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_sum, prediction.TRUE.), 
             na.action = na.omit) +
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(crs = w_asia_albers) +
  labs(fill = "Richness", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(face = "italic")
  )
```

@@

* Most cereal and legume crops predicted to be Levantine
* Cereals show diverging southern/northern ranges (legumes not so much)
* Not rye (Anatolia), chickpea (N. Mesopotamia)

```{r fig.height=10, fig-palaeodist-barely-pistachio-levant, eval=FALSE, cache=TRUE}
#| fig-cap: Contraction of reconstructed range of of wild barleys and pistachio in the Levant
palaeodist_hordeum <- do.call(c, c(
  filter(flora, str_starts(taxon, "Hordeum") | taxon == "Pistacia atlantica")$palaeodist,
  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum") | taxon == "Pistacia atlantica")$taxon))
))

@@

Wild barley (*Hordeum spontaneum*), its relatives, and pistachio (*Pistacia atlantica*) show a contraction of their predicted ranges from the Pleistocene to the Holocene,
concurrent with them being brought into cultivation [@fig-palaeodist-barely-pistachio-levant].
These two species also see marked declines in the archaeobotanical record from the Early PPNA/Early PPNB (where they were amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end.
We hypothesise that climate-linked range shifts were a factor driving these changes in the apparent economic important of different crops:
the decreasing availability of barley and pistachio in the terminal Pleistocene may have prompted cultivation as a strategy to retain access to them,
though in the long run it made them less attractive as staple crops compared to the more resilient wheat.
However, these dynamics are not seen in the majority of crop progenitors.

The wheat story.

<!-- TODO: figure with flax, Pistacia, Bolboschoenus -->

Flax has a very restricted distribution (consistent with low occurrence in founders paper?).
As does Pistacia atlantica, Bolboschoenus maritimus

<!-- TODO: figure with rye -->

*Secale cereale* = an Anatolian boy

@@

<!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
Is consistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.

>>>>

* [x] Stop modelling
- [ ] First draft:
  - [x] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [x] ENM
  - [x] Methods & Materials
  - [x] Results
  - [x] Discussion
  - [ ] Conclusion
- [ ] Figures
  - [x] fig-region
  - [x] fig-occ
  - [ ] tbl-occ-count (needs pdf cleanup)
  - [ ] tbl-predictors
- [ ] Outstanding TODOs
- [ ] Add missing references
- [ ] Copyedit
- [x] Appendix with all hindcast predictions (HTML?)
- [ ] Clean up package code
- [ ] Final proofread

@@

Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
This model can then be used to predict the range of the organism in question either in the same or a different environment.
@CITE suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.

@@

The major practical limitation of the hindcasting approach is that it relies on spatially explicit, high resolution palaeoenvironmental surfaces with continuous coverage of the region and periods of interest.
Until recently, this has not been widely available for most applications, which is perhaps why only a minority of studies use it [cf. @YaworskyEtAl2023].
In this study, we are able to take advantage of the increasing availability of high resolution, global palaeoclimate data derived from simulation experiments with general circulation models of climate [@BrownEtAl2018; @BrownEtAl2020; @KargerEtAl2023].

@@

```{r niche-predict}
# NOTE: high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  flora <- flora |>
    mutate(
      pred = map(niche, collect_predictions),
      palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = environment, .progress = TRUE),
      # Binarize
      palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
      palaeodist = map(palaeodist, mutate, absent = prediction.FALSE. > 0.5)
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))

# TODO write TIFs
```

@@

@@

```{r fig-palaeodist-sum}
#| fig-cap: Predicted species richness (sum of predicted ranges) by period
# TODO: expand period facet labels
ggplot() +
  facet_wrap(vars(factor(period, levels = climate_periods$code, labels = climate_periods$label))) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_sum, present), 
             na.action = na.omit) +
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(crs = w_asia_albers) +
  labs(fill = "Richness", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(face = "italic")
  )
```

@@

```{r fig-palaeodist-chickpea-rye}
#| fig-cap: Predicted palaeodistribution of wild chickpea and rye in the Early Holocene (11.7–8.2 ka)
chickpea_rye <- c("Secale cereale", "Cicer reticulatum")
palaeodist_chickpea_rye <- do.call(c, c(
  filter(flora, taxon %in% chickpea_rye)$palaeodist,
  list(along = list(taxon = filter(flora, taxon %in% chickpea_rye)$taxon))
))

ggplot() +
  facet_wrap(vars(taxon)) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_chickpea_rye, prediction.TRUE.) |>
               filter(period == "eh"), 
             na.action = na.omit) +
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(crs = w_asia_albers) +
  labs(fill = "P(present)", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "italic")
  )
```

Almost all the cereal and legume crop progenitors we modelled are predicted to have only been found in the Levant during the terminal Pleistocene and Early Holocene (see appendix).
Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-chickpea-rye). 
Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian [@fig-palaeodist-chickpea-rye].
This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].

```{r fig-palaeodist-locals}
#| fig-cap: Predicted palaeodistribution of flax, pistachio and clubrush in the Early Holocene (11.7–8.2 ka)
locals <- c("Linum bienne", "Pistacia atlantica", "Bolboschoenus maritimus")
palaeodist_locals <- do.call(c, c(
  filter(flora, taxon %in% locals)$palaeodist,
  list(along = list(taxon = filter(flora, taxon %in% locals)$taxon))
))

ggplot() +
  facet_wrap(vars(taxon), nrow = 2) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_locals, prediction.TRUE.) |>
               filter(period == "eh"), 
             na.action = na.omit) +
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(crs = w_asia_albers) +
  labs(fill = "P(present)", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "italic")
  )
```

Flax (*Linum bienne*) is predicted to have had a highly concentrated distribution in Cyprus and along the Mediterranean coast of the southern Levant.
This is consistent with its low ubiquity in archaeobotanical assemblages [@ArranzOtaeguiRoe2023], despite conventionally being considered a 'founder crop', and presumably implies that its domestication was similarly geographically constrained.
It is the only unambiguous crop progenitor with such a restricted range, though pistachio (*Pistacia atlantica*) and clubrush (*Bolboschoenus maritimus*) are similarly constrained to the Mediterranean coast (and North Africa, in the case of pistachio).

```{r fig.height=10, fig-palaeodist-barely-levant, eval=FALSE, cache=TRUE}
#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
palaeodist_hordeum <- do.call(c, c(
  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
))

@@

Wild barley (*Hordeum spontaneum*) its relatives show a contraction of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation [@fig-palaeodist-barely-levant].
It also sees a marked decline in the archaeobotanical record from the Early PPNA/Early PPNB (where it was amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
Pistachio (*Pistacia atlantica*) shows similar trends, but it is less certain that this species was cultivated in the Neolithic.
Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end.
We hypothesise that climate-linked range shifts were a factor driving these changes in the apparent economic important of different crops:
the decreasing availability of barley and pistachio in the terminal Pleistocene may have prompted cultivation as a strategy to retain access to them,
though in the long run it made them less attractive as staple crops compared to the more resilient wheat.
However, these dynamics are not seen in the majority of crop progenitors.

```{r fig-palaeodist-bread}
#| fig-cap: Combined predicted palaeodistributions of bread wheat progenitors
bread <- c("Aegilops tauschii", "Aegilops speltoides", 
           "Triticum turgidum dicoccum", "Triticum urartu")
palaeodist_bread <- do.call(c, c(
  filter(flora, taxon %in% bread)$palaeodist,
  list(along = list(taxon = filter(flora, taxon %in% bread)$taxon))
)) |>
  st_apply(c("x", "y", "period"), sum, na.rm = FALSE) |>
  mutate(present = factor(present))

ggplot() +
  facet_wrap(vars(fct_relabel(period, toupper))) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_bread, present), 
             na.action = na.omit) +
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_fill_batlow(reverse = TRUE, discrete = TRUE) +
  coord_sf(crs = w_asia_albers) +
  labs(fill = "N present", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(face = "italic")
  )
```

Bread wheat (*Triticum aestivum*), the most common wheat cultivar today, has a complex ancestry that involves two recent hybridisation events [@LevyFeldman2022]:
most recently between emmer (*Triticum turgidum dicoccum*) and a goatgrass (*Aegilops tauschii*) c. 9 ka, and before that, in emmer, between wild red einkorn (*T. urartu*) and another goatgrass (*Aegilops speltoides*).
Although there is fairly wide overlap between these species today, according to our modelling the only place all four (or even just three) are predicted to have coincided or been found in close proximity to eachother in the time frame of domestication is a narrow area around the Orontes river in the northern Levant [@fig-palaeodist-bread].
This area is c. 200 km from the only two sites in our archaeobotanical database where either both species of wheat or both species of wheat and the goatgrasses are found together: Tell Abu Hureyra and El Kowm II [@ArranzOtaeguiRoe2023].^[We are grateful to Benjamin Nowak for pointing this out to us.]
Taken together this suggests an origin of domesticated bread wheat, otherwise only loosely geographically constrained to the Levant–Upper Euphrates corridor [@LevyFeldman2022], is likely to occur in this vicinity.

@@

Is consistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
<!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->

### 2025-01-08 analysis/paper.tex
Add crop progenitors discussion

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Biogeography
    \item[$\boxtimes$]
      ENM
    \end{itemize}
  \item[$\boxtimes$]
    Methods \& Materials
  \item[$\boxtimes$]
    Results
  \item[$\square$]
    Discussion

    \begin{itemize}
    \tightlist
    \item[$\boxtimes$]
      General trends
    \item[$\square$]
      Case studies
    \item[$\boxtimes$]
      Comparison with archaeobot
    \end{itemize}
  \item[$\square$]
    Conclusion
  \end{itemize}
\item[$\square$]
  Figures

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    fig-region
  \item[$\boxtimes$]
    fig-occ
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-model-metrics
  \end{itemize}
\item[$\square$]
  Outstanding TODOs
\item[$\square$]
  Add missing references
\item[$\square$]
  Copyedit
\item[$\boxtimes$]
  Appendix with all hindcast predictions (HTML?)
\item[$\square$]
  Clean up package code
\item[$\square$]
  Final proofread
\end{itemize}

@@

Ecological niche modelling (ENM) or species distribution modelling (SDM)
is widely used by ecologists to predict the geographic range of a
species based on a set of environmental predictors. Essentially, it
involves combining records of where an organism has been observed with
environmental data (climate, topography, etc.) for those locations to
model the range of environmental values at which that species -- its
environmental niche. This model can then be used to predict the range of
the organism in question either in the same or a different environment.
\citet{CITE} suggests reserving the term `species distribution
modelling' for when the method is used to recover the verifiable range
of a species in a real and existing environment, and using `ecological
niche modelling' as the broader term covering hypothetical or predictive
applications -- a convention we follow here when referring to predictive
or `hindcast' models of past ranges. Within this overarching framework,
ecological niche modelling encompasses a wide range of applications and
a variety of potential environmental predictors, modelling approaches,
and methodologies, which we will not attempt to review here.

@@

\begin{itemize}
\tightlist
\item
  Most cereal and legume crops predicted to be Levantine
\item
  Cereals show diverging southern/northern ranges (legumes not so much)
\item
  Not rye (Anatolia), chickpea (N. Mesopotamia)
\end{itemize}

Wild barley (\emph{Hordeum spontaneum}), its relatives, and pistachio
(\emph{Pistacia atlantica}) show a contraction of their predicted ranges
from the Pleistocene to the Holocene, concurrent with them being brought
into cultivation \textbf{?@fig-palaeodist-barely-pistachio-levant}.
These two species also see marked declines in the archaeobotanical
record from the Early PPNA/Early PPNB (where they were amongst the most
common taxa) to the Late PPNB and Late Neolithic
\citep{ArranzOtaeguiRoe2023}. Conversely, the various wild wheat species
native to West Asia show almost no response to Pleistocene/Holocene
climate change, even within the Levant, and in the archaeobotanical
record wheat displays the opposite trend to barley and pistachio --
becoming gradually more abundant through the course of the Neolithic and
dominant by its end. We hypothesise that climate-linked range shifts
were a factor driving these changes in the apparent economic important
of different crops: the decreasing availability of barley and pistachio
in the terminal Pleistocene may have prompted cultivation as a strategy
to retain access to them, though in the long run it made them less
attractive as staple crops compared to the more resilient wheat.
However, these dynamics are not seen in the majority of crop
progenitors.

The wheat story.

Flax has a very restricted distribution (consistent with low occurrence
in founders paper?). As does Pistacia atlantica, Bolboschoenus maritimus

\emph{Secale cereale} = an Anatolian boy

>>>>

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Biogeography
    \item[$\boxtimes$]
      ENM
    \end{itemize}
  \item[$\boxtimes$]
    Methods \& Materials
  \item[$\boxtimes$]
    Results
  \item[$\boxtimes$]
    Discussion
  \item[$\square$]
    Conclusion
  \end{itemize}
\item[$\square$]
  Figures

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    fig-region
  \item[$\boxtimes$]
    fig-occ
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \end{itemize}
\item[$\square$]
  Outstanding TODOs
\item[$\square$]
  Add missing references
\item[$\square$]
  Copyedit
\item[$\boxtimes$]
  Appendix with all hindcast predictions (HTML?)
\item[$\square$]
  Clean up package code
\item[$\square$]
  Final proofread
\end{itemize}

@@

Ecological niche modelling or species distribution modelling is widely
used by ecologists to predict the geographic range of a species based on
a set of environmental predictors \citep{FranklinMiller2009}.
Essentially, it involves combining records of where an organism has been
observed with environmental data (climate, topography, etc.) for those
locations to model the range of environmental values at which that
species -- its environmental niche. This model can then be used to
predict the range of the organism in question either in the same or a
different environment. \citet{CITE} suggests reserving the term `species
distribution modelling' for when the method is used to recover the
verifiable range of a species in a real and existing environment, and
using `ecological niche modelling' as the broader term covering
hypothetical or predictive applications -- a convention we follow here
when referring to predictive or `hindcast' models of past ranges. Within
this overarching framework, ecological niche modelling encompasses a
wide range of applications and a variety of potential environmental
predictors, modelling approaches, and methodologies, which we will not
attempt to review here.

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-chickpea-rye-1.pdf}}

\caption{\label{fig-palaeodist-chickpea-rye}Predicted palaeodistribution
of wild chickpea and rye in the Early Holocene (11.7--8.2 ka)}

Almost all the cereal and legume crop progenitors we modelled are
predicted to have only been found in the Levant during the terminal
Pleistocene and Early Holocene (see appendix). Part of this may be do
with the fact that both our initial flora and training occurrence
dataset have a strong bias towards the southern Levant, but strikingly
the modelled \emph{current} ranges of these plants do tend to include
Anatolia and the Zagros, so this cannot be the only factor. One notable
exception is the wild ancestor of chickpea (\emph{Cicer reticulatum}),
which is predicted to have a distribution centered on Northern
Mesopotamia but encompassing much of the the `hilly flanks' (except the
southern Levant, \textbf{?@fig-chickpea-rye}). Another is rye
(\emph{Secale cereale}), which is inferred to be primarily Western
Anatolian Figure~\ref{fig-palaeodist-chickpea-rye}. This is perhaps
relevant to rye's unusual domestication history, as a crop of West Asian
origin that was apparently not first cultivated until much later than
the `founder crops', in Europe \citep{SchreiberEtAl2021}.

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-locals-1.pdf}}

\caption{\label{fig-palaeodist-locals}Predicted palaeodistribution of
flax, pistachio and clubrush in the Early Holocene (11.7--8.2 ka)}

Flax (\emph{Linum bienne}) is predicted to have had a highly
concentrated distribution in Cyprus and along the Mediterranean coast of
the southern Levant. This is consistent with its low ubiquity in
archaeobotanical assemblages \citep{ArranzOtaeguiRoe2023}, despite
conventionally being considered a `founder crop', and presumably implies
that its domestication was similarly geographically constrained. It is
the only unambiguous crop progenitor with such a restricted range,
though pistachio (\emph{Pistacia atlantica}) and clubrush
(\emph{Bolboschoenus maritimus}) are similarly constrained to the
Mediterranean coast (and North Africa, in the case of pistachio).

Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation
\textbf{?@fig-palaeodist-barely-levant}. It also sees a marked decline
in the archaeobotanical record from the Early PPNA/Early PPNB (where it
was amongst the most common taxa) to the Late PPNB and Late Neolithic
\citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia atlantica})
shows similar trends, but it is less certain that this species was
cultivated in the Neolithic. Conversely, the various wild wheat species
native to West Asia show almost no response to Pleistocene/Holocene
climate change, even within the Levant, and in the archaeobotanical
record wheat displays the opposite trend to barley and pistachio --
becoming gradually more abundant through the course of the Neolithic and
dominant by its end. We hypothesise that climate-linked range shifts
were a factor driving these changes in the apparent economic important
of different crops: the decreasing availability of barley and pistachio
in the terminal Pleistocene may have prompted cultivation as a strategy
to retain access to them, though in the long run it made them less
attractive as staple crops compared to the more resilient wheat.
However, these dynamics are not seen in the majority of crop
progenitors.

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-bread-1.pdf}}

\caption{\label{fig-palaeodist-bread}Combined predicted
palaeodistributions of bread wheat progenitors}

Bread wheat (\emph{Triticum aestivum}), the most common wheat cultivar
today, has a complex ancestry that involves two recent hybridisation
events \citep{LevyFeldman2022}: most recently between emmer
(\emph{Triticum turgidum dicoccum}) and a goatgrass (\emph{Aegilops
tauschii}) c.~9 ka, and before that, in emmer, between wild red einkorn
(\emph{T. urartu}) and another goatgrass (\emph{Aegilops speltoides}).
Although there is fairly wide overlap between these species today,
according to our modelling the only place all four (or even just three)
are predicted to have coincided or been found in close proximity to
eachother in the time frame of domestication is a narrow area around the
Orontes river in the northern Levant Figure~\ref{fig-palaeodist-bread}.
This area is c.~200 km from the only two sites in our archaeobotanical
database where either both species of wheat or both species of wheat and
the goatgrasses are found together: Tell Abu Hureyra and El Kowm II
\citep{ArranzOtaeguiRoe2023}.\footnote{We are grateful to Benjamin Nowak
  for pointing this out to us.} Taken together this suggests an origin
of domesticated bread wheat, otherwise only loosely geographically
constrained to the Levant--Upper Euphrates corridor
\citep{LevyFeldman2022}, is likely to occur in this vicinity.
